---
title: '03 - Module selection, functional & TF enrichment, figures (Fig 4-6; Suppl. Tables 4-6)'
author: "Cesare Marino"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    keep_md: true
  pdf_document:
    toc: true
---

```{r global setup, include=FALSE}
# ========== GLOBAL SETTINGS ==========

knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE, 
  fig.width = 12, 
  fig.height = 8,
  cache = FALSE,
  error = FALSE,
  eval = TRUE
)

options(repos = c(CRAN = "https://cran.rstudio.com/"))
options(width = 100)
options(dplyr.summarise.inform = FALSE)


# ========== GLOBAL OUTPUT SETTINGS ==========

here::i_am("code/03_analysis_and_figures.Rmd")  # path to THIS Rmd from project root

params <- list(
  save_outputs = TRUE,
  results_dir = here::here("results", format(Sys.Date()))
)

if (params$save_outputs && !dir.exists(params$results_dir)) {
  dir.create(params$results_dir, recursive = TRUE)
}

# ========== FILE PATHS ==========

wgcna_file  <- here::here("data", "new_wgcna.csv")
trrust_file <- here::here("data", "trrust_rawdata.mouse.tsv")


# ========== ANALYSIS PARAMETERS ==========

egs_threshold_percentile <- 0.85
min_gene_count <- 10
min_conditions <- 2

# ========== CONDITION SETS ==========

condition_sets <- list(
  early    = c("PKD_wk2", "PKD_wk3", "PKD_wk6"),
  advanced = c("PKD_Moderate", "PKD_Advanced"),
  combined = c("PKD_wk2", "PKD_wk3", "PKD_wk6", "PKD_Moderate", "PKD_Advanced")
)


# Set this to one of: "early", "advanced", "combined" or custom
downstream_condition_set <- "early"


# ========== UTILITY FUNCTIONS ==========

clean_wgcna_data <- function(df, keep_conditions) {
  df %>%
    filter(experiment %in% keep_conditions) %>%
    filter(!is.na(module)) %>%
    transmute(
      condition = experiment,
      gene_id = MOUSE_GENE_SYMBOL,
      entrez_id = MOUSE_GENE_NCBI_GENE_ID,
      log2fc = log2fc,
      pvalue = pvalue,
      padj = padj,
      module = module,
      egs_score = eg_score
    )
}


analyze_modules <- function(df, egs_percentile, min_genes, min_conds) {
    egs_raw <- df %>%
    select(module, condition, egs_score) %>%
    distinct()
  
  if (nrow(egs_raw) > 0) {
    egs_matrix <- egs_raw %>%
      pivot_wider(names_from = condition, values_from = egs_score, values_fill = NA)
  
    # Only rename if there are columns besides 'module'
    if (ncol(egs_matrix) > 1) {
      egs_matrix <- egs_matrix %>%
        rename_with(~ paste0("EGS_", .), .cols = -module)
    }
  } else {
    egs_matrix <- tibble(module = character(0))
  }
  
  stats <- df %>%
    group_by(module) %>%
    summarise(
      gene_list = paste(unique(gene_id), collapse = ";"),
      gene_count = n_distinct(gene_id),
      conds_present = n_distinct(condition),
      max_abs_egs = max(abs(egs_score), na.rm = TRUE),
      mean_abs_egs = mean(egs_score, na.rm = TRUE),
      sd_egs = sd(egs_score, na.rm = TRUE),
      mean_egs = mean(egs_score, na.rm = TRUE),
      direction = ifelse(mean_egs > 0, "Upregulated", "Downregulated"),
      .groups = "drop"
    ) %>%
    left_join(egs_matrix, by = "module") %>%
    mutate(
      threshold_value_egs = quantile(max_abs_egs, egs_percentile, na.rm = TRUE),
      passes_egs = max_abs_egs >= threshold_value_egs,
      passes_size = gene_count >= min_genes,
      passes_conditions = conds_present >= min_conds,
      selected = passes_egs & passes_size & passes_conditions
    ) %>%
    arrange(desc(max_abs_egs))
  
  return(stats)
}

run_module_analysis <- function(condition_vector, input_data) {
  df <- clean_wgcna_data(input_data, condition_vector)
  stats <- analyze_modules(df, egs_threshold_percentile, min_gene_count, min_conditions)
  sel_ids <- stats %>% filter(selected) %>% pull(module)

  list(
    data_filtered = df,
    modules_stats = stats,
    modules_sel_ids = sel_ids,
    modules_selected = stats %>%
      filter(selected) %>%
      select(module, direction, gene_count, max_abs_egs, mean_egs),
    modules_genes = stats %>%
      filter(selected) %>%
      mutate(
        genes = strsplit(gene_list, ";"),
        gene_count_check = gene_count
      ) %>%
      select(module, genes, gene_count_check)
  )
}

```

## Step 1: Module Selection and Eigengene Activity Analysis

This step loads the WGCNA results and filters modules based on condition-specific eigengene scores (EGS), building the `results` object (early, advanced, combined selection sets) used by every downstream figure and table. Module selection uses the 85th-percentile EGS threshold with the minimum gene-count and condition filters; the selection thresholds are printed for the record. Figures are built in the dedicated Figure sections below.


```{r egs_heatmaps_full, message=FALSE, warning=FALSE}
# =============================================================================
# FULL EGS HEATMAPS
# =============================================================================

library(pheatmap)
library(RColorBrewer)
library(grid)
library(dplyr)
library(tidyr)
library(tibble)

# — Load the new CSV 
new_wgcna <- read_csv(wgcna_file)

stopifnot(all(c("experiment","module","eg_score","gene_symbol") %in% colnames(new_wgcna)))


# Run module selection for each group
results <- list(
  early    = run_module_analysis(condition_sets$early,    new_wgcna),
  advanced = run_module_analysis(condition_sets$advanced, new_wgcna)
)


merge_results <- function(res1, res2) {
  list(
    modules_sel_ids = union(res1$modules_sel_ids, res2$modules_sel_ids),

    modules_selected = bind_rows(res1$modules_selected, res2$modules_selected) %>%
      distinct(module, .keep_all = TRUE),

    modules_stats = bind_rows(res1$modules_stats, res2$modules_stats) %>%
      distinct(module, .keep_all = TRUE),

    data_filtered = bind_rows(
      res1$data_filtered,
      res2$data_filtered
    ),

    modules_genes = {
      gene_col <- setdiff(names(res1$modules_genes), "module")[1]
      bind_rows(res1$modules_genes, res2$modules_genes) %>%
        distinct(module, .data[[gene_col]], .keep_all = TRUE)
    }
  )
}

results$combined <- merge_results(results$early, results$advanced)


# Use the long-format EGS table computed in Script 3
# Columns: condition, module, egs_score, etc.
new_wgcna <- results$combined$data_filtered

# --- EGS selection thresholds (85th percentile), kept for the record ---
cat("egs_threshold_percentile =", egs_threshold_percentile, "\n")
cat("EGS cutoff (early)   :", unique(results$early$modules_stats$threshold_value_egs), "\n")
cat("EGS cutoff (advanced):", unique(results$advanced$modules_stats$threshold_value_egs), "\n")
```

## Figure 4: Module EGS heatmap and early-module top genes

Figure 4 is built as two panels, each computed once and exported as a vector PDF;
the two panels are combined manually afterwards. Panel (a) is the combined-selection
module EGS heatmap with Early/Severe row blocks (`p_top`). Panel (b) is the top-5 genes
per early-selected module across all conditions, colour-clamped to [-3, 3] and row-linked
to the Early block of panel (a).

### Fig 4a: module EGS heatmap (Early vs Severe)

```{r fig4a_module_egs_heatmap, message=FALSE, warning=FALSE, fig.width=10, fig.height=12}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(pheatmap)
  library(grid)
  library(gridExtra)
  library(gtable)
})

# =============================================================================
# SETTINGS
# =============================================================================
desired_order <- c("PKD_wk2","PKD_wk3","PKD_wk6","PKD_Moderate","PKD_Advanced")
pal_bwr <- grDevices::colorRampPalette(c("#4575b4","white","#d73027"))(100)

stage_cols <- c("Severe"="#C77700","Early"="#1F78B4")
annotation_colors_mod <- list(
  stage = stage_cols,
  direction = c("Upregulated"="#d73027","Downregulated"="#4575b4"),
  activity_level = c("Very High"="#8B0000","High"="#CD5C5C","Moderate"="#F0E68C","Lower"="#D3D3D3"),
  size_category = c("Large"="#006400","Medium"="#32CD32","Small"="#90EE90")
)

# =============================================================================
# TOP PANEL: MODULE EGS HEATMAP (Early vs Severe blocks)
# =============================================================================
new_wgcna <- results$combined$data_filtered

mods_early    <- results$early$modules_sel_ids
mods_advanced <- results$advanced$modules_sel_ids
mods_severe   <- setdiff(mods_advanced, mods_early)
mods_all      <- results$combined$modules_sel_ids

mat_mod <- new_wgcna %>%
  filter(module %in% mods_all, condition %in% desired_order) %>%
  mutate(condition = factor(condition, levels = desired_order)) %>%
  select(module, condition, egs_score) %>%
  distinct() %>%
  pivot_wider(names_from = condition, values_from = egs_score) %>%
  column_to_rownames("module") %>%
  as.matrix()

mat_mod <- mat_mod[, desired_order, drop = FALSE]
mat_mod <- mat_mod[rowSums(!is.na(mat_mod)) > 0, , drop = FALSE]

ann_mod <- results$combined$modules_selected %>%
  filter(module %in% rownames(mat_mod)) %>%
  mutate(
    activity_level = case_when(
      max_abs_egs >= quantile(max_abs_egs, 0.8, na.rm = TRUE) ~ "Very High",
      max_abs_egs >= quantile(max_abs_egs, 0.6, na.rm = TRUE) ~ "High",
      max_abs_egs >= quantile(max_abs_egs, 0.4, na.rm = TRUE) ~ "Moderate",
      TRUE ~ "Lower"
    ),
    size_category = case_when(
      gene_count >= 50 ~ "Large",
      gene_count >= 25 ~ "Medium",
      TRUE ~ "Small"
    ),
    stage = case_when(
      module %in% mods_severe ~ "Severe",
      module %in% mods_early  ~ "Early",
      TRUE ~ NA_character_
    )
  ) %>%
  column_to_rownames("module")

ann_mod <- ann_mod[rownames(mat_mod), , drop = FALSE]

# Order rows as: Severe Up, Severe Down, Early Up, Early Down, within each by mean EGS
df_ord <- tibble(module = rownames(mat_mod)) %>%
  left_join(ann_mod %>% as.data.frame() %>% rownames_to_column("module"), by = "module") %>%
  mutate(
    block = case_when(
      stage=="Severe" & direction=="Upregulated"   ~ 1L,
      stage=="Severe" & direction=="Downregulated" ~ 2L,
      stage=="Early"  & direction=="Upregulated"   ~ 3L,
      stage=="Early"  & direction=="Downregulated" ~ 4L,
      TRUE ~ 99L
    ),
    row_mean = rowMeans(mat_mod[module, , drop = FALSE], na.rm = TRUE)
  ) %>%
  arrange(block, desc(row_mean), module)

mat_mod <- mat_mod[df_ord$module, , drop = FALSE]
ann_mod <- ann_mod[df_ord$module, , drop = FALSE]

n_su <- sum(ann_mod$stage=="Severe" & ann_mod$direction=="Upregulated", na.rm = TRUE)
n_sd <- sum(ann_mod$stage=="Severe" & ann_mod$direction=="Downregulated", na.rm = TRUE)
n_eu <- sum(ann_mod$stage=="Early"  & ann_mod$direction=="Upregulated", na.rm = TRUE)
gaps_mod <- c(n_su, n_su + n_sd, n_su + n_sd + n_eu)
gaps_mod <- gaps_mod[gaps_mod > 0 & gaps_mod < nrow(mat_mod)]
if (length(gaps_mod) == 0) gaps_mod <- NULL

bk_mod <- seq(-2, 2, length.out = 101)

p_top <- pheatmap(
  mat_mod,
  color = pal_bwr,
  breaks = bk_mod,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  gaps_row = gaps_mod,
  annotation_row = ann_mod[, c("stage","direction","activity_level","size_category"), drop = FALSE],
  annotation_colors = annotation_colors_mod,
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize = 8,
  fontsize_row = 7,
  border_color = "grey70",
  main = "Combined modules - all conditions (Early vs Severe)",
  silent = TRUE
)

# Color rownames by stage (as you wanted)
stage_rowname_col <- ifelse(ann_mod$stage == "Severe", stage_cols[["Severe"]], stage_cols[["Early"]])
idx_rn <- which(grepl("row_names", p_top$gtable$layout$name) | grepl("rownames", p_top$gtable$layout$name))
if (length(idx_rn) >= 1) {
  idx_rn <- idx_rn[1]
  rn_grob <- p_top$gtable$grobs[[idx_rn]]
  if (inherits(rn_grob, "text")) {
    rn_grob$gp$col <- stage_rowname_col
    rn_grob$gp$fontface <- "bold"
    p_top$gtable$grobs[[idx_rn]] <- rn_grob
  } else if (!is.null(rn_grob$children) && length(rn_grob$children) > 0) {
    for (k in seq_along(rn_grob$children)) {
      if (inherits(rn_grob$children[[k]], "text")) {
        rn_grob$children[[k]]$gp$col <- stage_rowname_col
        rn_grob$children[[k]]$gp$fontface <- "bold"
      }
    }
    p_top$gtable$grobs[[idx_rn]] <- rn_grob
  }
}

# =============================================================================
# EXPORT PANEL (a): module EGS heatmap only
# =============================================================================
# =============================================================================
# EXPORT UPPER PANEL ONLY (Module EGS heatmap)
# =============================================================================

stopifnot(exists("p_top"), exists("params"))

out_dir <- file.path(params$results_dir, "Composite_figures")
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

# ---- Vector PDF (BEST for Illustrator) ----
out_pdf <- file.path(out_dir, "UpperPanel_ModuleEGS_heatmap_AI.pdf")

grDevices::pdf(out_pdf,
               width = 8,      # adjust if needed
               height = 10,    # adjust if needed
               useDingbats = FALSE)

grid::grid.newpage()
grid::grid.draw(p_top$gtable)

grDevices::dev.off()

cat("Saved upper panel PDF to:\n",
    normalizePath(out_pdf, winslash = "/"), "\n")


# ---- Optional high-resolution TIFF fallback ----
out_tif <- file.path(out_dir, "UpperPanel_ModuleEGS_heatmap_AI.tiff")

grDevices::tiff(out_tif,
                width = 8,
                height = 10,
                units = "in",
                res = 600,
                compression = "lzw")

grid::grid.newpage()
grid::grid.draw(p_top$gtable)

grDevices::dev.off()

cat("Saved upper panel TIFF to:\n",
    normalizePath(out_tif, winslash = "/"), "\n")
```

### Fig 4b: early-module top-5 genes (all conditions, linked to panel a)

```{r fig4b_early_top5_genes, message=FALSE, warning=FALSE, fig.width=14, fig.height=6}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(pheatmap)
  library(grid)
  library(gridExtra)
})

stopifnot(exists("results"), exists("condition_sets"), exists("params"))
stopifnot(all(c("early","combined") %in% names(results)))

# IMPORTANT: this is what keeps the lower panel linked to the upper
stopifnot(exists("ann_mod"))

# Same column order as upper heatmap
desired_order <- c("PKD_wk2","PKD_wk3","PKD_wk6","PKD_Moderate","PKD_Advanced")

pal_bwr <- grDevices::colorRampPalette(c("#4575b4","white","#d73027"))(100)
k_per_module <- 5

# =============================================================================
# Pick top-5 genes per EARLY module (rank by max abs log2fc across EARLY conditions)
# =============================================================================
gene2mod <- results$early$modules_genes %>%
  filter(module %in% results$early$modules_sel_ids) %>%
  select(module, genes) %>%
  tidyr::unnest_longer(genes) %>%
  rename(gene_id = genes) %>%
  distinct(gene_id, module)

df_early <- results$early$data_filtered %>%
  filter(gene_id %in% gene2mod$gene_id) %>%
  select(gene_id, condition, log2fc) %>%
  distinct()

mat_early <- df_early %>%
  pivot_wider(names_from = condition, values_from = log2fc) %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

score_tbl <- tibble(
  gene_id = rownames(mat_early),
  score = apply(mat_early, 1, function(x) max(abs(x), na.rm = TRUE))
)

rank_tbl <- gene2mod %>%
  inner_join(score_tbl, by = "gene_id") %>%
  group_by(module) %>%
  arrange(desc(score), .by_group = TRUE) %>%
  slice_head(n = k_per_module) %>%
  ungroup()

genes_keep <- unique(rank_tbl$gene_id)

# =============================================================================
# Build gene x ALL conditions matrix from COMBINED object, and FORCE SAME col order
# =============================================================================
df_all <- results$combined$data_filtered %>%
  filter(gene_id %in% genes_keep, condition %in% desired_order) %>%
  mutate(condition = factor(condition, levels = desired_order)) %>%
  select(gene_id, condition, log2fc) %>%
  distinct()

mat_all <- df_all %>%
  pivot_wider(names_from = condition, values_from = log2fc) %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

mat_all <- mat_all[, desired_order, drop = FALSE]
mat_all <- mat_all[rowSums(!is.na(mat_all)) > 0, , drop = FALSE]

gene2mod_for_plot <- rank_tbl %>% select(gene_id, module) %>% distinct()
gene2mod_for_plot <- gene2mod_for_plot[match(rownames(mat_all), gene2mod_for_plot$gene_id), , drop = FALSE]

keep_ok <- !is.na(gene2mod_for_plot$module)
mat_all <- mat_all[keep_ok, , drop = FALSE]
gene2mod_for_plot <- gene2mod_for_plot[keep_ok, , drop = FALSE]

# =============================================================================
# Module order linked to the upper heatmap EARLY section (top to bottom)
# =============================================================================
early_order_top <- rownames(ann_mod)[ann_mod$stage == "Early"]
mod_present <- unique(gene2mod_for_plot$module)

mod_levels <- early_order_top[early_order_top %in% mod_present]
mod_levels_extra <- setdiff(mod_present, mod_levels)
if (length(mod_levels_extra) > 0) mod_levels <- c(mod_levels, mod_levels_extra)

gene2mod_for_plot$module <- factor(gene2mod_for_plot$module, levels = mod_levels)

module_cols <- setNames(grDevices::hcl.colors(length(mod_levels), palette = "Dynamic"), mod_levels)

# =============================================================================
# Group genes by module; cluster within module
# =============================================================================
mat_list <- vector("list", length(mod_levels))
names(mat_list) <- mod_levels

for (m in mod_levels) {
  genes_m <- gene2mod_for_plot$gene_id[gene2mod_for_plot$module == m]
  sub <- mat_all[genes_m, , drop = FALSE]
  if (nrow(sub) >= 2) {
    hc <- stats::hclust(stats::dist(sub))
    sub <- sub[hc$order, , drop = FALSE]
  }
  mat_list[[m]] <- sub
}
mat_grouped <- do.call(rbind, mat_list)

ann_row_gene <- data.frame(
  module = factor(
    gene2mod_for_plot$module[match(rownames(mat_grouped), gene2mod_for_plot$gene_id)],
    levels = mod_levels
  ),
  stringsAsFactors = FALSE
)
rownames(ann_row_gene) <- rownames(mat_grouped)

# ONLY requested change: clamp DISPLAY range to -3..3
bk_gene <- seq(-3, 3, length.out = 101)

# =============================================================================
# Split into 4 panels (contiguous module blocks)
# =============================================================================
n_panels <- 4
panel_assign <- split(mod_levels, cut(seq_along(mod_levels), breaks = n_panels, labels = FALSE))

make_gene_panel <- function(mod_subset, show_numeric_legend = FALSE) {
  idx <- which(as.character(ann_row_gene$module) %in% mod_subset)
  sub_mat <- mat_grouped[idx, , drop = FALSE]
  sub_ann <- ann_row_gene[idx, , drop = FALSE]

  block_sizes <- vapply(mod_subset, function(mm) sum(as.character(sub_ann$module) == mm), integer(1))
  gaps <- cumsum(block_sizes)
  gaps <- gaps[gaps < nrow(sub_mat)]

  pheatmap(
    sub_mat,
    color = pal_bwr,
    breaks = bk_gene,
    cluster_cols = FALSE,
    cluster_rows = FALSE,
    gaps_row = gaps,
    annotation_row = sub_ann,
    annotation_colors = list(module = module_cols),
    show_rownames = TRUE,
    show_colnames = TRUE,
    fontsize_row = 7,
    fontsize_col = 9,
    angle_col = 90,          # straight column names
    cellwidth = 18,
    cellheight = 11,
    border_color = NA,
    legend = show_numeric_legend,
    annotation_legend = FALSE,
    annotation_names_row = FALSE,
    silent = TRUE
  )$gtable
}

gene_gt <- lapply(seq_along(panel_assign), function(i) {
  make_gene_panel(panel_assign[[i]], show_numeric_legend = (i == length(panel_assign)))
})

bottom_row <- gridExtra::arrangeGrob(grobs = gene_gt, ncol = 4, widths = rep(1, 4))

grid::grid.newpage()
grid::grid.draw(bottom_row)

if (isTRUE(params$save_outputs)) {
  out_dir <- file.path(params$results_dir, "Composite_figures")
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

  out_pdf <- file.path(out_dir, "LowerPanel_EARLYtop5genes_allConditions_clamp-3to3_LINKED.pdf")
  grDevices::pdf(out_pdf, width = 14, height = 6, useDingbats = FALSE)
  grid::grid.newpage(); grid::grid.draw(bottom_row)
  grDevices::dev.off()

  cat("Saved lower panel PDF to:\n", normalizePath(out_pdf, winslash = "/"), "\n")
}
```



## Functional Enrichment Analysis

This step performs pathway enrichment analysis on selected modules using enrichR, filtering for statistically significant terms across multiple databases (GO, KEGG, WikiPathways, MSigDB). Outputs include enrichment tables, barplots for each module, and heatmaps combining module activity with functional annotations.

```{r step2-functional-enrichment, fig.cap="Functional Enrichment Results"}

# =============================================================================
# STEP 2: FUNCTIONAL ENRICHMENT ANALYSIS
# =============================================================================

# ——— Enrichment parameters ———
min_module_genes       <- 10
min_pathway_size       <- 5
max_pathway_size       <- 500
min_intersection_ratio <- 0.2
pvalue_threshold       <- 0.05

setEnrichrSite("Enrichr")

# Select databases
preferred_databases <- c(
  "GO_Biological_Process_2025",
  "KEGG_2026",
  "WikiPathways_2024_Mouse",
  "Reactome_Pathways_2024",
  "MSigDB_Hallmark_2020"
)

available_databases <- intersect(preferred_databases,
                                 listEnrichrDbs()$libraryName)
stopifnot(length(available_databases) > 0)

# ——— Fixed color mapping (publication-consistent) ———
db_colors <- c(
  GO_Biological_Process_2025 = "#0072B2",  # blue
  KEGG_2026                  = "#E69F00",  # orange
  MSigDB_Hallmark_2020       = "#009E73",  # green
  Reactome_Pathways_2024     = "#D55E00",  # vermillion
  WikiPathways_2024_Mouse    = "#CC79A7"   # purple
)

# ——— Function definition ———
perform_module_enrichment <- function(gene_list, module_name, databases) {

  if (length(gene_list) < min_module_genes) return(NULL)

  tryCatch({

    enr <- enrichr(gene_list, databases)
    out <- data.frame()

    for (db in names(enr)) {

      df <- enr[[db]]
      if (is.null(df) || nrow(df) == 0) next

      df2 <- df %>%
        mutate(
          module                    = module_name,
          database                  = db,
          gene_count                = length(gene_list),
          intersection_size         = sapply(
                                        Genes,
                                        function(x) length(strsplit(x, ";")[[1]])
                                      ),
          min_intersection_required = ceiling(
                                        length(gene_list) *
                                        min_intersection_ratio
                                      )
        ) %>%
        filter(
          intersection_size >= min_pathway_size,
          intersection_size <= max_pathway_size,
          intersection_size >= min_intersection_required,
          P.value < pvalue_threshold
        ) %>%
        select(
          module, database, gene_count, Term, P.value,
          Adjusted.P.value, Combined.Score,
          Genes, intersection_size, min_intersection_required
        )

      if (nrow(df2) > 0) out <- bind_rows(out, df2)
    }

    if (nrow(out) == 0) NULL else out

  }, error = function(e) NULL)
}

# ——— Prepare output directory ———
union_out_dir <- file.path(params$results_dir, "enrichment_union")
if (!dir.exists(union_out_dir))
  dir.create(union_out_dir, recursive = TRUE)

# ——— Build union of module gene lists ———
union_modules <- bind_rows(
  results$early$modules_genes,
  results$advanced$modules_genes,
  results$combined$modules_genes
) %>%
  distinct(module, .keep_all = TRUE)

# ——— Run enrichment ———
all_enrichments <- data.frame()
enrichment_list <- list()

for (i in seq_len(nrow(union_modules))) {

  mod   <- union_modules$module[i]
  genes <- union_modules$genes[[i]]

  res <- perform_module_enrichment(
    genes,
    mod,
    available_databases
  )

  if (!is.null(res)) {
    enrichment_list[[mod]] <- res
    all_enrichments        <- bind_rows(all_enrichments, res)
  }
}

# ——— Annotation ———
annotations_union <- union_modules %>%
  select(module, gene_count = gene_count_check) %>%
  left_join(
    all_enrichments %>%
      group_by(module) %>%
      slice_min(P.value, n = 1) %>%
      transmute(
        module,
        top_term              = Term,
        best_pvalue           = P.value,
        top_intersection_size = intersection_size,
        top_database          = database
      ) %>%
      mutate(
        has_enrichment     = "Yes",
        significance_level = case_when(
          best_pvalue < 0.001 ~ "Very High",
          best_pvalue < 0.01  ~ "High",
          best_pvalue < 0.05  ~ "Moderate",
          TRUE                ~ "Lower"
        )
      ),
    by = "module"
  ) %>%
  replace_na(list(
    has_enrichment        = "No",
    significance_level    = "None",
    top_intersection_size = 0
  ))

results$union_enrichment <- list(
  enrichments = enrichment_list,
  annotations = annotations_union,
  plots       = list()
)

# =============================================================================
# Generate & Save Per-Module Plots
# =============================================================================

for (mod in names(enrichment_list)) {

  df <- enrichment_list[[mod]]

  # ---- Top 12 Overall ----
  top_overall <- df %>%
    arrange(desc(Combined.Score)) %>%
    slice_head(n = 12) %>%
    mutate(
      neg_log_p = -log10(P.value),
      term_short = ifelse(
        nchar(Term) > 50,
        paste0(substr(Term, 1, 47), "..."),
        Term
      )
    )

  p1 <- ggplot(
    top_overall,
    aes(reorder(term_short, neg_log_p),
        neg_log_p,
        fill = database)
  ) +
    geom_col(color = "black", linewidth = 0.3) +
    scale_fill_manual(values = db_colors, drop = FALSE) +
    coord_flip() +
    theme_minimal() +
    labs(title = paste("Top12 Overall —", mod),
         x = NULL,
         y = "-log10(p)")

  # ---- Top 3 Per DB ----
  top_db <- df %>%
    group_by(database) %>%
    arrange(P.value) %>%
    slice_head(n = 3) %>%
    ungroup() %>%
    mutate(
      neg_log_p = -log10(P.value),
      term_short = ifelse(
        nchar(Term) > 50,
        paste0(substr(Term, 1, 47), "..."),
        Term
      )
    )

  p2 <- ggplot(
    top_db,
    aes(reorder(term_short, neg_log_p),
        neg_log_p,
        fill = database)
  ) +
    geom_col(color = "black", linewidth = 0.3) +
    scale_fill_manual(values = db_colors, drop = FALSE) +
    coord_flip() +
    theme_minimal() +
    labs(title = "Top3 per DB",
         x = NULL,
         y = "-log10(p)")

  # ---- DB Summary ----
  db_sum <- df %>%
    group_by(database) %>%
    summarise(terms = n(), .groups = "drop")

  p3 <- ggplot(
    db_sum,
    aes(database, terms, fill = database)
  ) +
    geom_col(color = "black", linewidth = 0.3) +
    scale_fill_manual(values = db_colors, drop = FALSE) +
    theme_minimal() +
    labs(title = "DB Summary",
         x = NULL,
         y = "Count") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none")

  combo <- p1 / p2 / p3

  results$union_enrichment$plots[[mod]] <- combo

  ggsave(
    file.path(union_out_dir,
              paste0("enrichment_", mod, ".png")),
    combo,
    width = 8,
    height = 12,
    dpi = 150
  )
}

# =============================================================================
# FIGURE 5 — Top 10 Overall Pathways (6 Modules)
# =============================================================================

modules_of_interest <- c(
  "rKIDNEY_33",
  "rKIDNEY_28m",
  "rKIDNEY_64m",
  "rKIDNEY_29",
  "rKIDNEY_96m",
  "rKIDNEY_124"
)

plot_top10_overall <- function(df, module_name) {

  df_top10 <- df %>%
    arrange(desc(Combined.Score)) %>%
    slice_head(n = 10) %>%
    mutate(
      neg_log_p = -log10(P.value),
      term_short = ifelse(
        nchar(Term) > 60,
        paste0(substr(Term, 1, 57), "..."),
        Term
      )
    )

  ggplot(
    df_top10,
    aes(reorder(term_short, neg_log_p),
        neg_log_p,
        fill = database)
  ) +
    geom_col(color = "black", linewidth = 0.4) +
    scale_fill_manual(values = db_colors, drop = FALSE) +
    coord_flip() +
    theme_minimal(base_size = 15) +
    theme(
      legend.position = "none",
      plot.title = element_text(size = 18, face = "bold")
    ) +
    labs(title = module_name,
         x = NULL,
         y = "-log10(p)")
}

plot_list <- lapply(modules_of_interest, function(mod) {
  df <- results$union_enrichment$enrichments[[mod]]
  if (!is.null(df) && nrow(df) > 0)
    plot_top10_overall(df, mod)
  else
    ggplot() + theme_void() + labs(title = paste(mod, "(no enrichment)"))
})

names(plot_list) <- modules_of_interest

final_plot <- (
  plot_list[[1]] | plot_list[[2]]
) / (
  plot_list[[3]] | plot_list[[4]]
) / (
  plot_list[[5]] | plot_list[[6]]
)

ggsave(
  file.path(union_out_dir,
            "Top10_overall_6modules_NOLEGEND.png"),
  final_plot,
  width = 14,
  height = 18,
  dpi = 300
)

```



## TF Enrichment Analysis

This step performs transcription factor enrichment analysis using TRRUST database, identifying TFs significantly enriched in each module through hypergeometric testing. Outputs include TF enrichment tables, frequency plots, static and interactive bipartite networks showing TF-module regulatory relationships, and hub TF identification.

```{r step3-tf-enrichment, fig.cap="TF Enrichment Analysis Results"}
# =============================================================================
# STEP 3: TF ENRICHMENT ANALYSIS (MAIN AND COMBINED)
# =============================================================================

# === PARAMETERS ===
min_tf_targets            <- 1
min_intersection_required <- 2
min_fold_enrichment       <- 1.5
fdr_threshold             <- 0.10

# === BACKGROUND ===
background_genes <- unique(results$combined$data_filtered$gene_id)

# === MODULE SELECTION ===
selected_modules <- results$combined$modules_sel_ids
modules_genes <- results$combined$modules_genes %>%
  filter(module %in% selected_modules)

# === TRRUST FILTER ===
tf_regulons <- read_tsv(trrust_file, col_names = c("TF", "Target", "Mode", "PMID"), show_col_types = FALSE)
tf_regulons <- tf_regulons %>%
  filter(Target %in% background_genes)

# === FUNCTION ===
perform_tf_enrichment <- function(module_genes, tf_regulons, background_genes) {
  results <- list()
  for (tf in unique(tf_regulons$TF)) {
    targets <- tf_regulons %>% filter(TF == tf) %>% pull(Target) %>% unique()
    if (length(targets) < min_tf_targets) next

    overlap <- intersect(module_genes, targets)
    if (length(overlap) < min_intersection_required) next

    m <- length(intersect(targets, background_genes))
    N <- length(background_genes)
    k <- length(module_genes)
    x <- length(overlap)

    pval     <- phyper(x - 1, m, N - m, k, lower.tail = FALSE)
    expected <- (k * m) / N
    fold     <- x / max(expected, 1)
    if (fold < min_fold_enrichment) next

    modes <- tf_regulons %>% filter(TF == tf, Target %in% overlap) %>% pull(Mode)
    dommode <- if (all(c("Activation", "Repression") %in% modes)) "Mixed"
      else if ("Activation" %in% modes) "Activation"
      else if ("Repression" %in% modes) "Repression"
      else "Unknown"

    pmids <- tf_regulons %>% filter(TF == tf, Target %in% overlap) %>% pull(PMID) %>% unique()

    results[[tf]] <- data.frame(
      TF                = tf,
      overlap_count     = x,
      module_size       = k,
      tf_targets_total  = m,
      expected_overlap  = expected,
      fold_enrichment   = fold,
      pvalue            = pval,
      dominant_mode     = dommode,
      pmid_count        = length(pmids),
      overlapping_genes = paste(overlap, collapse = ";"),
      stringsAsFactors  = FALSE
    )
  }
  if (length(results) == 0) return(NULL)
  df <- bind_rows(results)
  df$fdr <- p.adjust(df$pvalue, method = "fdr")
  df <- df %>%
    filter(fdr <= fdr_threshold)
  return(df)
}

# === RUN ===
tf_enrich_combined <- list()

for (mod in selected_modules) {
  genes <- modules_genes %>% filter(module == mod) %>% pull(genes) %>% .[[1]]
  df <- perform_tf_enrichment(genes, tf_regulons, background_genes)
  if (!is.null(df)) {
    tf_enrich_combined[[mod]] <- df %>% mutate(module = mod)
  }
}

# === STORE ===
results$combined$tf_enrichments <- tf_enrich_combined

# === FLATTEN + FILTER ===
tf_df <- bind_rows(tf_enrich_combined)
significant_tf_enrichments <- tf_df %>%
  filter(fdr <= fdr_threshold, overlap_count >= min_intersection_required)

# === NETWORK DATA ===
max_modules <- 12
max_tfs_per_module <- 8

top_modules <- significant_tf_enrichments %>%
  count(module, name = "n_tfs") %>%
  arrange(desc(n_tfs)) %>%
  slice_head(n = max_modules) %>%
  pull(module)

network_data <- significant_tf_enrichments %>%
  filter(module %in% top_modules) %>%
  group_by(module) %>%
  arrange(desc(fold_enrichment)) %>%
  slice_head(n = max_tfs_per_module) %>%
  ungroup() %>%
  transmute(
    from         = TF,
    to           = module,
    weight       = fold_enrichment,
    edge_mode    = dominant_mode,
    module_label = sub(".*:", "", module)
  )

tf_module_membership <- expand.grid(
  TF     = unique(network_data$from),
  module = modules_genes$module,
  stringsAsFactors = FALSE
) %>%
  rowwise() %>%
  mutate(
    gene_list = list(modules_genes$genes[modules_genes$module == module][[1]]),
    is_member = TF %in% gene_list[[1]]
  ) %>%
  ungroup() %>%
  filter(is_member) %>%
  group_by(TF) %>%
  summarise(
    member_modules = paste(gsub(".*[:|]", "", module), collapse = ","),
    .groups = "drop"
  )


# === BUILD GRAPH ===
g <- igraph::graph_from_data_frame(network_data, directed = FALSE)
igraph::V(g)$type <- igraph::V(g)$name %in% network_data$from
igraph::V(g)$label <- ifelse(
  igraph::V(g)$type,
  igraph::V(g)$name,
  network_data$module_label[match(igraph::V(g)$name, network_data$to)]
)
igraph::V(g)$member_modules <- ifelse(
  igraph::V(g)$type & igraph::V(g)$name %in% tf_module_membership$TF,
  tf_module_membership$member_modules[match(igraph::V(g)$name, tf_module_membership$TF)],
  ""
)
igraph::V(g)$fill <- ifelse(
  igraph::V(g)$type,
  "grey80",
  ifelse(
    results$combined$modules_selected$direction[match(igraph::V(g)$name, results$combined$modules_selected$module)] == "Upregulated",
    "#d73027", "#4575b4"
  )
)
igraph::E(g)$edge_mode <- network_data$edge_mode
igraph::E(g)$weight <- network_data$weight

# === LAYOUT ===
layout_mat <- igraph::layout_as_bipartite(g, types = igraph::V(g)$type)
tfi  <- which(igraph::V(g)$type)
modi <- which(!igraph::V(g)$type)
layout_mat[tfi,1]  <- seq(-5, 5, length.out = length(tfi))
layout_mat[modi,1] <- seq(-4, 4, length.out = length(modi))
layout_mat[tfi,2]  <-  1
layout_mat[modi,2] <- -1

# === PLOT ===


p <- ggraph(g, layout = "manual", x = layout_mat[,1], y = layout_mat[,2]) +
  geom_edge_link(aes(edge_colour = edge_mode, edge_width = weight), alpha = 0.8) +
  geom_node_point(aes(fill = fill), shape = 21, size = 5, colour = "black") +
  geom_node_text(aes(label = label), repel = TRUE, size = 3) +
  geom_node_text(aes(label = ifelse(member_modules != "", member_modules, "")),
                 size = 2.5, colour = "white", fontface = "bold") +
  scale_edge_color_manual(
    values = c(
      Activation = "#d73027",
      Repression = "#4575b4",
      Mixed = "#8B008B",
      Unknown = "grey60"
    ),
    name = "Regulation"
  ) +
  scale_edge_width(name = "Fold Enrichment") +
  scale_fill_identity() +
  theme_void() +
  ggtitle("TF–Module Bipartite Network (Combined Selection)")

# === SAVE PLOT ===
ggsave(
  file.path(params$results_dir, "tf_module_bipartite_combined.png"),
  plot = p,
  width = 10,
  height = 7,
  dpi = 300
)



# === PREPARE MAIN TABLE: one row per TF–Module edge in network ===
main_table <- network_data %>%
  rename(TF = from, Module = to, Fold_Enrichment = weight, Edge_Mode = edge_mode) %>%
  mutate(Module_Label = sub(".*:", "", Module)) %>%
  # TRUE TF–module membership calculated by unnesting module gene lists
  left_join(
    modules_genes %>%
      unnest_longer(genes) %>%
      rename(TF = genes, Module_Label = module) %>%
      mutate(Is_TF_Module_Member = TRUE),
    by = c("TF", "Module_Label")
  ) %>%
  mutate(Is_TF_Module_Member = replace_na(Is_TF_Module_Member, FALSE)) %>%
  select(TF, Module, Module_Label, Fold_Enrichment, Edge_Mode, Is_TF_Module_Member) %>%
  arrange(Module_Label, desc(Fold_Enrichment))

# === ADD EXTRA COLUMNS from enrichment stats ===
df_enrich <- bind_rows(tf_enrich_combined, .id = "Module") %>%
  mutate(Module_Label = sub(".*:", "", Module)) %>%
  select(
    Module_Label, TF,
    Overlap_Count = overlap_count,
    TF_Targets_Total = tf_targets_total,
    Expected_Overlap = expected_overlap,
    Pvalue = pvalue,
    FDR = fdr,
    Dominant_Mode = dominant_mode,
    Overlapping_Genes = overlapping_genes
  )

# === MERGE into one master table ===
final_table <- main_table %>%
  left_join(df_enrich, by = c("Module_Label", "TF")) %>%
  select(
    TF,
    Module_Label,
    Fold_Enrichment,
    Overlap_Count,
    TF_Targets_Total,
    Expected_Overlap,
    FDR,
    Edge_Mode,
    Dominant_Mode,
    Overlapping_Genes,
    Is_TF_Module_Member
  )

# === WRITE TO ONE EXCEL FILE ===
wb <- openxlsx::createWorkbook()
openxlsx::addWorksheet(wb, "TF_Module_Bipartite_Network")
openxlsx::writeData(wb, "TF_Module_Bipartite_Network", final_table)

out_path <- file.path(params$results_dir, "tf_module_bipartite_summary.xlsx")
openxlsx::saveWorkbook(wb, out_path, overwrite = TRUE)


```

###fig 6 

```{r chord_export_both, message=FALSE, warning=FALSE}

library(circlize)
library(dplyr)

# --------------------------------------------------
# PREPARE chord_df WITH ALL TFs (no filtering)
# --------------------------------------------------
chord_df <- network_data %>%
  mutate(
    TF     = from,
    Module = to,
    Weight = weight,
    Mode   = edge_mode
  ) %>%
  select(TF, Module, Weight, Mode)

TFs     <- unique(chord_df$TF)
Modules <- unique(chord_df$Module)

# Module colors
module_cols_raw <- ifelse(
  results$combined$modules_selected$direction[
    match(Modules, results$combined$modules_selected$module)
  ] == "Upregulated",
  "#FF0000",   # red
  "#0072B2"    # blue
)
names(module_cols_raw) <- Modules

# TF colors
tf_cols_raw <- rep("#BBBBBB", length(TFs))
names(tf_cols_raw) <- TFs

group_colors <- c(tf_cols_raw, module_cols_raw)

# Edge colors (soft)
edge_color_map <- c(
  Activation = "#FF0000",
  Repression = "#0072B2",
  Mixed      = "#800080",
  Unknown    = "#777777"
)
chord_df$col <- adjustcolor(edge_color_map[chord_df$Mode], alpha.f = 0.30)



# ==================================================
# FUNCTION: Draw chord diagram with no labels
# ==================================================
draw_chord_nolabels <- function() {
  circos.clear()
  circos.par(start.degree=90, gap.degree=2, track.margin=c(0.02,0.02))

  chordDiagram(
    chord_df[,c("TF","Module","Weight","col")],
    grid.col = group_colors,
    transparency = 0.70,
    directional = 0,
    annotationTrack = "grid",
    annotationTrackHeight = mm_h(4),
    link.sort = TRUE,
    link.largest.ontop = TRUE,
    col = chord_df$col
  )

  circos.track(
    track.index=1,
    ylim=c(0,1),
    bg.border=NA,
    panel.fun=function(x,y){}
  )
}

# ==================================================
# FUNCTION: Draw chord diagram with ALL RADIAL LABELS
# ==================================================
draw_chord_radial <- function() {
  circos.clear()
  circos.par(start.degree=90, gap.degree=2, track.margin=c(0.02,0.02))

  chordDiagram(
    chord_df[,c("TF","Module","Weight","col")],
    grid.col = group_colors,
    transparency = 0.65,
    directional = 0,
    annotationTrack = "grid",
    annotationTrackHeight = mm_h(4),
    link.sort = TRUE,
    link.largest.ontop = TRUE,
    col = chord_df$col
  )

  # Replace labels with radial labels for ALL sectors
  circos.trackPlotRegion(
    track.index=1,
    ylim=c(0,1),
    bg.border = NA,
    panel.fun = function(x,y){
      circos.text(
        CELL_META$xcenter,
        CELL_META$ylim[1] - 0.30,
        CELL_META$sector.index,
        facing="clockwise",
        niceFacing=TRUE,
        adj=c(0.5,-0.2),
        cex=1.2,
        font=2
      )
    }
  )
}

# ==================================================
# EXPORT BOTH VERSIONS
# ==================================================

# Output paths
png_radial <- file.path(params$results_dir, "TF_Module_Chord_Radial.png")
png_nolabel <- file.path(params$results_dir, "TF_Module_Chord_NoLabels.png")

svg_radial <- file.path(params$results_dir, "TF_Module_Chord_Radial.svg")
svg_nolabel <- file.path(params$results_dir, "TF_Module_Chord_NoLabels.svg")

eps_radial <- file.path(params$results_dir, "TF_Module_Chord_Radial.eps")
eps_nolabel <- file.path(params$results_dir, "TF_Module_Chord_NoLabels.eps")


# PNG export
png(png_radial, width=2800, height=2800, res=300)
draw_chord_radial()
dev.off()

png(png_nolabel, width=2800, height=2800, res=300)
draw_chord_nolabels()
dev.off()

# SVG export
svg(svg_radial, width=12, height=12)
draw_chord_radial()
dev.off()

svg(svg_nolabel, width=12, height=12)
draw_chord_nolabels()
dev.off()

# EPS export
setEPS()
postscript(eps_radial, width=12, height=12)
draw_chord_radial()
dev.off()

setEPS()
postscript(eps_nolabel, width=12, height=12)
draw_chord_nolabels()
dev.off()

cat("Export complete:\n",
    png_radial, "\n",
    png_nolabel, "\n",
    svg_radial, "\n",
    svg_nolabel, "\n",
    eps_radial, "\n",
    eps_nolabel, "\n")

```

```{r sensitivity-percentile}

# Sensitivity analysis: key module retention at 80th, 85th, 90th percentile
key_modules <- c("rKIDNEY_33", "rKIDNEY_124", "rKIDNEY_28m", "rKIDNEY_29", "rKIDNEY_64m", "rKIDNEY_103")

new_wgcna_raw <- read_csv(wgcna_file)

sensitivity_results <- lapply(c(0.80, 0.85, 0.90), function(pct) {
  df <- clean_wgcna_data(new_wgcna_raw, condition_sets$early)
  stats <- analyze_modules(df, pct, min_gene_count, min_conditions)
  sel <- stats %>% filter(selected) %>% pull(module)
  
  data.frame(
    percentile = pct,
    n_modules_selected = length(sel),
    key_modules_retained = paste(key_modules[key_modules %in% sel], collapse = "; "),
    all_key_retained = all(key_modules %in% sel)
  )
})

sensitivity_df <- bind_rows(sensitivity_results)
print(sensitivity_df)
```

